Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-09-11 Documents/M1/realizableFermionicM1/sections/appendix.tex
minor edits

For the analysis of constraint-preserving IMEX schemes, additional coefficients are defined \cite{hu_etal_2018} (cf. Eq.~\eqref{eq:imexStagesRewrite}).  
First, let
\begin{equation}
  b_{ii} = \f{1}{a_{ii}}, \quad
  b_{ij} = -\f{1}{a_{ii}}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad
  \tilde{b}_{ij} = -\f{1}{a_{ii}}\Big(\tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}\Big).  
\end{equation}
Then, for IMEX schemes of Type~A \cite{dimarcoPareschi2013},
\begin{equation}
  \begin{aligned}
    c_{i0} &= 1-\sum_{j=1}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
    c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj}, \\
    \tilde{c}_{i0} &= 0, \quad &
    \tilde{c}_{ij} &= \tilde{a}_{ij} + \sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}.
  \end{aligned}
  \label{eq:positivityCoefficientsA}
\end{equation}
while for IMEX schemes of Type~ARS \cite{ascher_etal_1997}
\begin{equation}
  \begin{aligned}
    c_{i0} &= 1-\sum_{j=2}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
    c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj} \\
    \tilde{c}_{i0} &= \tilde{a}_{i1}+\sum_{j=2}^{i-1}a_{ij}\tilde{b}_{j1}, \quad &
    \tilde{c}_{ij} &= \tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}.  
  \end{aligned}
  \label{eq:positivityCoefficientsARS}
\end{equation}

@@

Here we prove that a PD-IMEX scheme  of type~A (i.e., conforming to Definition~\ref{def:PD-IMEX} with Eq.~\eqref{eq:positivityConditionsTypeA} replacing Eq.~\eqref{eq:positivityConditionsTypeARS} in item 3) does not exist.  
First, for a three-stage, GSA IMEX scheme of type~A the double Butcher tableau is
\begin{equation*}
  \begin{array}{c | c c c}
  	      0       & 0              & 0              & 0 \\
  	\tilde{c}_{2} & \tilde{a}_{21} & 0              & 0 \\
  	\tilde{c}_{3} & \tilde{a}_{31} & \tilde{a}_{32} & 0 \\ \hline
  	              & \tilde{a}_{31} & \tilde{a}_{32} & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
    c_{1} & a_{11} & 0      & 0      \\
  	c_{2} & a_{21} & a_{22} & 0      \\
  	c_{3} & a_{31} & a_{32} & a_{33} \\ \hline
  	      & a_{31} & a_{32} & a_{33}
  \end{array}.
\end{equation*}
First we consider the equality constraints.  
Consistency of the implicit coefficients and second-order accuracy in the streaming limit (Eqs.~\eqref{eq:implicitConsistency} and \eqref{eq:orderConditionsEx}, respectively) give
\begin{align}
  a_{31} + a_{32} + a_{33} = 1, \quad \tilde{a}_{31} + \tilde{a}_{32} = 1, \quad \text{and}\quad \tilde{a}_{32}\,\tilde{a}_{21} =  \f{1}{2}.
  \label{eq:1stOrderGeneral}
\end{align}
Accuracy in the diffusion limit (Eq.~\eqref{eq:diffusionCondition}) requires 
\begin{align}
  \tilde{a}_{21} = a_{22}
  \quad\text{and}\quad 
  -\f{a_{32}\,\tilde{a}_{21}}{a_{22}\,a_{33}} + \f{1}{a_{33}} = 1.
\label{eq:diffusionAccurate}
\end{align}
The first constraint in Eq.~\eqref{eq:1stOrderGeneral} together with the constraints in Eq.~\eqref{eq:diffusionAccurate} gives
\begin{align}
  a_{31} = 0.
  \label{eq:a31is0}
\end{align}
Next we consider the inequality constraints.  
The constraint-preserving property in Eq.~\eqref{eq:positivityConditionsTypeA} requires $a_{11}, a_{22}, a_{33}>0$, and 
\begin{equation}
  c_{21} = \frac{a_{21}}{a_{11}} \geq 0, \quad
  c_{31} = \frac{a_{31}}{a_{11}} - \frac{a_{32}\,a_{21}}{a_{22}\,a_{11}} \geq 0, \quad\text{and}\quad
  c_{32} = \frac{a_{32}}{a_{22}} \geq 0.
  \label{eq:constraintsTypeA}
\end{equation}
Since $a_{31} = 0$, we have
\begin{equation*}
  c_{31} = - \frac{a_{32}a_{21}}{a_{22}a_{11}}.  
\end{equation*}
Then, since $a_{21} \geq 0$, $a_{32} \geq 0$, $a_{22} > 0$ and $a_{11} >0$, we have $c_{31} \leq 0$, and the inequality constraints in Eq.~\eqref{eq:constraintsTypeA} hold only for $c_{31} = 0$.
Unfortunately, $c_{31} = 0$ gives $c_{\mbox{\tiny Sch}} = \min \Big\{\,\f{c_{21}}{\tilde{c}_{21}}, \f{c_{31}}{\tilde{c}_{31}}, \f{c_{32}}{\tilde{c}_{32}}\Big\} = 0$.
Therefore, a three-stage PD-IMEX scheme (Definition~\ref{def:PD-IMEX}) of type~A does not exist.

>>>>

For the analysis of constraint-preserving IMEX schemes, additional coefficients are defined \cite{hu_etal_2018} (cf. Eq.~\eqref{eq:imexStagesRewrite}).  
First, let
\begin{equation}
  b_{ii} = \f{1}{a_{ii}}, \quad
  b_{ij} = -\f{1}{a_{ii}}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad
  \tilde{b}_{ij} = -\f{1}{a_{ii}}\Big(\tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}\Big).  
\end{equation}
Then, for IMEX schemes of Type~A \cite{dimarcoPareschi2013},
\begin{equation}
  \begin{aligned}
    c_{i0} &= 1-\sum_{j=1}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
    c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj}, \\
    \tilde{c}_{i0} &= 0, \quad &
    \tilde{c}_{ij} &= \tilde{a}_{ij} + \sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}.
  \end{aligned}
  \label{eq:positivityCoefficientsA}
\end{equation}
while for IMEX schemes of Type~ARS \cite{ascher_etal_1997}
\begin{equation}
  \begin{aligned}
    c_{i0} &= 1-\sum_{j=2}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
    c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj} \\
    \tilde{c}_{i0} &= \tilde{a}_{i1}+\sum_{j=2}^{i-1}a_{ij}\tilde{b}_{j1}, \quad &
    \tilde{c}_{ij} &= \tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}.  
  \end{aligned}
  \label{eq:positivityCoefficientsARS}
\end{equation}
(Note that $c_{i1}=\tilde{c}_{i1}=0$ in Eq.~\eqref{eq:positivityCoefficientsARS} so that $\sum_{j=0}^{i-1}c_{ij}=1$.)

@@

Here we prove that a PD-IMEX scheme  of type~A (i.e., conforming to Definition~\ref{def:PD-IMEX}, but with Eq.~\eqref{eq:positivityConditionsTypeA} replacing Eq.~\eqref{eq:positivityConditionsTypeARS} in item 3) does not exist.  
First, for a three-stage, GSA IMEX scheme of type~A the double Butcher tableau is
\begin{equation*}
  \begin{array}{c | c c c}
  	      0       & 0              & 0              & 0 \\
  	\tilde{c}_{2} & \tilde{a}_{21} & 0              & 0 \\
  	\tilde{c}_{3} & \tilde{a}_{31} & \tilde{a}_{32} & 0 \\ \hline
  	              & \tilde{a}_{31} & \tilde{a}_{32} & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
    c_{1} & a_{11} & 0      & 0      \\
  	c_{2} & a_{21} & a_{22} & 0      \\
  	c_{3} & a_{31} & a_{32} & a_{33} \\ \hline
  	      & a_{31} & a_{32} & a_{33}
  \end{array}.
\end{equation*}
First we consider the equality constraints.  
Consistency of the implicit coefficients and second-order accuracy in the streaming limit (Eqs.~\eqref{eq:implicitConsistency} and \eqref{eq:orderConditionsEx}, respectively) give
\begin{align}
  a_{31} + a_{32} + a_{33} = 1, \quad \tilde{a}_{31} + \tilde{a}_{32} = 1, \quad \text{and}\quad \tilde{a}_{32}\,\tilde{a}_{21} =  \f{1}{2}.
  \label{eq:1stOrderGeneral}
\end{align}
Accuracy in the diffusion limit (Eq.~\eqref{eq:diffusionCondition}) requires 
\begin{align}
  \tilde{a}_{21} = a_{22}
  \quad\text{and}\quad 
  -\f{a_{32}\,\tilde{a}_{21}}{a_{22}\,a_{33}} + \f{1}{a_{33}} = 1.
\label{eq:diffusionAccurate}
\end{align}
The first constraint in Eq.~\eqref{eq:1stOrderGeneral} together with the constraints in Eq.~\eqref{eq:diffusionAccurate} gives
\begin{align}
  a_{31} = 0.
  \label{eq:a31is0}
\end{align}
Next we consider the inequality constraints.  
The constraint-preserving property in Eq.~\eqref{eq:positivityConditionsTypeA} requires $a_{11}, a_{22}, a_{33}>0$, and 
\begin{equation}
  c_{21} = \frac{a_{21}}{a_{11}} \geq 0, \quad
  c_{31} = \frac{a_{31}}{a_{11}} - \frac{a_{32}\,a_{21}}{a_{22}\,a_{11}} \geq 0, \quad\text{and}\quad
  c_{32} = \frac{a_{32}}{a_{22}} \geq 0.
  \label{eq:constraintsTypeA}
\end{equation}
Since $a_{31} = 0$, we have
\begin{equation*}
  c_{31} = - \frac{a_{32}a_{21}}{a_{22}a_{11}}.  
\end{equation*}
Then, since $a_{21} \geq 0$, $a_{32} \geq 0$, $a_{22} > 0$ and $a_{11} >0$, we have $c_{31} \leq 0$, and the inequality constraints in Eq.~\eqref{eq:constraintsTypeA} hold only for $c_{31} = 0$.
Unfortunately, $c_{31} = 0$ gives $c_{\mbox{\tiny Sch}} = \min \Big\{\,\f{c_{21}}{\tilde{c}_{21}}, \f{c_{31}}{\tilde{c}_{31}}, \f{c_{32}}{\tilde{c}_{32}}\Big\} = 0$.
Therefore, a three-stage PD-IMEX scheme (Definition~\ref{def:PD-IMEX}) of type~A does not exist.

### 2018-09-11 Documents/M1/realizableFermionicM1/sections/imex.tex
minor edits

Unfortunately, the ``diffusion limit requirement" in Eq.~\eqref{eq:diffusionCondition}, together with the order conditions given by Eqs.~\eqref{orderConditions1} and \eqref{orderConditions2} (with $\alpha=0$), and the positivity conditions on $c_{ij}$ and $\tilde{c}_{ij}$, result in too many constraints to obtain a second-order accurate constraint-preserving IMEX scheme.
\footnote{It was shown in \cite{hu_etal_2018} --- without the diffusion limit requirement --- that the minimum number of stages for constraint-preserving IMEX schemes of type~ARS is four.)  
We are also concerned about increasing the number of stages, and thereby the number of implicit solves, since the implicit solve will dominate the computational cost of the IMEX scheme with more realistic collision operators (e.g., inelastic scattering).}  
To reduce the number of constraints, and accommodate accuracy in the diffusion limit, we relax the requirement of overall second-order accuracy of the IMEX scheme.  
Instead, we only require the scheme to be second-order accurate in the streaming limit ($\vect{\cQ}=0$).  
This gives the order conditions
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=1
  \quad\text{and}\quad
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2},
  \label{eq:orderConditionsEx}
\end{equation}
where the first condition (consistency condition) is required for first-order accuracy.  
We then seek to design IMEX schemes of Type~ARS conforming to the following working definition
\begin{define}
  Let PD-IMEX be a consistent, constraint-preserving, diffusion accurate, and low-stage GSA IMEX scheme satisfying the following properties
  \begin{enumerate}
    \item Consistency of the implicit coefficients
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
      \label{eq:implicitConsistency}
    \end{equation}
    \item Second-order in the streaming limit; i.e., satisfies Eq.~\eqref{eq:orderConditionsEx}.
    \item Constraint-preserving; i.e. satisfies Eq.~\eqref{eq:positivityConditionsTypeARS}, with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\mbox{\tiny Sch}}>0$.
    \item Well-behaved in the diffusion limit; i.e., satisfies Eq.~\eqref{eq:diffusionCondition}.
    \item Less than four stages ($s\le3$).
    \item Globally stiffly accurate (GSA) $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$.
  \end{enumerate}
  \label{def:PD-IMEX}
\end{define}
Fortunately, IMEX schemes of type~ARS satisfying these properties are easy to find, and we provide an example with $s=3$ (two implicit solves) in \ref{app:butcherTables} (scheme PD-ARS; see \ref{app:PD-ARS} for further details).  
In the streaming limit, this scheme is identical to the optimal second-order accurate SSP-RK method \cite{gottlieb_etal_2001}.  
It is also very similar to the scheme given in \cite{mcclarren_etal_2008} (see scheme PC2 in \ref{app:butcherTables}), which is also a GSA IMEX scheme of type ARS with $s=3$.  
Scheme PC2 is second-order in the streaming limit, has been demonstrated to work well in the diffusion limit \cite{mcclarren_etal_2008,radice_etal_2013}, and satisfies the positivity conditions in Eq.~\eqref{eq:positivityConditionsTypeARS}.  
However, $c_{\mbox{\tiny Sch}}=0$ (our primary motivation for finding an alternative).  
In Section~\ref{sec:numerical}, we show numerically that the accuracy of scheme PD-ARS is comparable to the accuracy of scheme PC2.

>>>>

Unfortunately, the ``diffusion limit requirement" in Eq.~\eqref{eq:diffusionCondition}, together with the order conditions given by Eqs.~\eqref{orderConditions1} and \eqref{orderConditions2} (with $\alpha=0$), and the positivity conditions on $c_{ij}$ and $\tilde{c}_{ij}$, result in too many constraints to obtain a second-order accurate constraint-preserving IMEX scheme.
\footnote{It was shown in \cite{hu_etal_2018} --- without the diffusion limit requirement --- that the minimum number of stages for constraint-preserving IMEX schemes of type~ARS is four.)  
We are also concerned about increasing the number of stages, and thereby the number of implicit solves, since the implicit solve will dominate the computational cost of the IMEX scheme with more realistic collision operators (e.g., inelastic scattering).}  
To reduce the number of constraints and accommodate accuracy in the diffusion limit, we relax the requirement of overall second-order accuracy of the IMEX scheme.  
Instead, we only require the scheme to be second-order accurate in the streaming limit ($\vect{\cQ}=0$).  
This gives the order conditions
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=1
  \quad\text{and}\quad
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2},
  \label{eq:orderConditionsEx}
\end{equation}
where the first condition (consistency condition) is required for first-order accuracy.  
We then seek to design IMEX schemes of Type~ARS conforming to the following working definition
\begin{define}
  Let PD-IMEX be a consistent, constraint-preserving, diffusion accurate, and low-stage GSA IMEX scheme satisfying the following properties
  \begin{enumerate}
    \item Consistency of the implicit coefficients
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
      \label{eq:implicitConsistency}
    \end{equation}
    \item Second-order in the streaming limit; i.e., satisfies Eq.~\eqref{eq:orderConditionsEx}.
    \item Constraint-preserving; i.e. satisfies Eq.~\eqref{eq:positivityConditionsTypeARS}, with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\mbox{\tiny Sch}}>0$.
    \item Well-behaved in the diffusion limit; i.e., satisfies Eq.~\eqref{eq:diffusionCondition}.
    \item Less than four stages ($s\le3$).
    \item Globally stiffly accurate (GSA) $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$.
  \end{enumerate}
  \label{def:PD-IMEX}
\end{define}
Fortunately, IMEX schemes of type~ARS satisfying these properties are easy to find, and we provide an example with $s=3$ (two implicit solves) in \ref{app:butcherTables} (scheme PD-ARS; see \ref{app:PD-ARS} for further details).  
In the streaming limit, this scheme is identical to the optimal second-order accurate SSP-RK method \cite{gottlieb_etal_2001}.  
It is also very similar to the scheme given in \cite{mcclarren_etal_2008} (see scheme PC2 in \ref{app:butcherTables}), which is also a GSA IMEX scheme of type ARS with $s=3$.  
Scheme PC2 is second-order in the streaming limit, has been demonstrated to work well in the diffusion limit \cite{mcclarren_etal_2008,radice_etal_2013}, and satisfies the positivity conditions in Eq.~\eqref{eq:positivityConditionsTypeARS}.  
However, $c_{\mbox{\tiny Sch}}=0$ (our primary motivation for finding an alternative).  
In Section~\ref{sec:numerical}, we show numerically that the accuracy of scheme PD-ARS is comparable to the accuracy of scheme PC2.
